---
title: "CSSS 554: Homework 4"
subtitle: "Department of Biostatistics @ University of Washington"
author:
- Alejandro Hernandez
date: "Winter Quarter 2025"
output: pdf_document
---

```{r setup, include=F}
# Clear environment
rm(list=ls())

# Setup options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results='hide')
options(knitr.kable.NA = '-', digits = 2)
options(survey.lonely.psu="adjust")
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "allcode")]

```

```{r load}
# Load relevant packages
library(dplyr)
library(haven)
library(geodata)
library(INLA)
library(sf)
library(sn)
library(SUMMER)
library(surveyPrev)

# Load data locally
indicator <- "unmet_family"
year <- 2018
country <- "Benin"

## Load data locally
file_path <- "../data/Benin20172018/BJBR71DT/BJBR71FL.DTA"  # data file
dhsData <- as.data.frame(haven::read_dta(file_path))
indicator <- "unmet_family"
data <- surveyPrev::getDHSindicator(dhsData, indicator = indicator)
head(data)  # View head

# Handle missing data (there is none)
anyNA(data)

```

```{r}
# Get admin 1 info
poly.adm1 <- geodata::gadm(country, level=1, path=tempdir())
poly.adm1 <- sf::st_as_sf(poly.adm1)
admin1.info <- surveyPrev::adminInfo(poly.adm=poly.adm1, admin=1, by.adm="NAME_1")
head(admin1.info$data)  # View head

# Get admin 2 info
poly.adm2 <- gadm(country, level=2, path=tempdir())
poly.adm2 <- sf::st_as_sf(poly.adm2)
admin2.info <- adminInfo(poly.adm=poly.adm2, admin=2, by.adm="NAME_2", by.adm.upper="NAME_1")
head(admin2.info$data)  # View head

# Get geographic data locally
sf_use_s2(FALSE)  # IDK why this works but it fixes an error
file_path_geo <- "../data/Benin20172018/BJGE71FL/BJGE71FL.shp" # shapefile
geo <- sf::st_read(file_path_geo)

# Get cluster information at Admin 1 and 2
cluster.info <- clusterInfo(geo, poly.adm1, poly.adm2)

```


```{r}
## Get direct estimates of Admin 1 prevalence
res_ad1 <- surveyPrev::directEST(data, cluster.info, admin = 1)
benin_ad1 <- res_ad1$res.admin1 %>%
  select("Admin 1 Name"=admin1.name, "Direct Estimate"=direct.est, 
         "Variance of Estimate"=direct.var) %>%
  mutate("SE of Estimate" = sqrt(.$"Variance of Estimate"))

## Get direct estimates of Admin 2 prevalence
res_ad2 <- surveyPrev::directEST(data, cluster.info, admin = 2)
benin_ad2 <- res_ad2$res.admin2 %>%
  select("Admin 2 Name"=admin2.name, "Direct Estimate"=direct.est, 
         "Variance of Estimate"=direct.var) %>%
  mutate("SE of Estimate" = sqrt(.$"Variance of Estimate"))

# Take a peek at estimates table
head(benin_ad1)
head(benin_ad2)

# Histogram of direct estimates
hist(benin_ad1$`Direct Estimate`, main="Admin 1")
hist(benin_ad2$`Direct Estimate`, main="Admin 2")

# Histogram of direct estimate variances
hist(benin_ad1$`Variance of Estimate`, main="Admin 1")
hist(benin_ad2$`Variance of Estimate`, main="Admin 2")

# Range of direct estimate variances
summary(benin_ad1$`Variance of Estimate`)
summary(benin_ad2$`Variance of Estimate`)

```


## Question 1

(a) Obtain Fay-Herriot estimates at Admin1. To describe the model, define the logit of the
weighted estimate, $p_i^w$, to be $\hat{θ_i} = log [\hat{p_i^w}/(1-\hat{p_i^w})]$ and $\hat{V_i}$ to be the estimated design-based variance of $\hat{θ_i}$. Fit the Fay-Herriot model,

$\hat{\theta_i}|\theta_i \sim N(\theta_i,\hat{V_i})$
$\hat{\theta_i} = \alpha + b_i$
$b_i \sim BYM2(\sigma_b^2,\phi)$

Then, plot $\widetilde{p_i}$ against $\widehat{p_i}^w$. Plot the posterior standard deviation from the Fay-Herriot model versus the standard errors of $\widehat{p_i}^w$. Comment on these two scatterplots.

```{r question-1a, fig.asp=1}
# Fit Fay-Herriot model of Admin 1 prevalence
benin_ad1_bym2 <- fhModel(data, cluster.info, admin.info=admin1.info, admin=1, 
                          model="bym2", aggregation=FALSE)
head(benin_ad1_bym2$res.admin1)

# Add estimates to Admin 1 table
benin_ad1 <- cbind(benin_ad1, 
                   benin_ad1_bym2$res.admin1 %>%
                     select("F-H Mean"=mean, "F-H Variance"=var, "F-H SD"=sd))

# Map Admin 1 posterior means
SUMMER::mapPlot(data=benin_ad1, geo=poly.adm1,
                by.data="Admin 1 Name", by.geo="NAME_1", variables="F-H Mean",
                size = 0.01, direction = -1)

# Assess whether posterior variances may be too small
summary(benin_ad1$`F-H Variance`)
benin_ad1_bym2$res.admin1[c(1,6,2,7)]
# The variances are not extremely small nor the CIs extremely narrow 

# Plot of direct estimates against posterior means 
plot(x=benin_ad1$'Direct Estimate', y=benin_ad1$'F-H Mean', 
     ylab="Posterior Proportion", xlab="Direct Estimate",
     ylim=c(0.2,.4), xlim=c(0.2,.4))
abline(0,1,col="red")

# Plot of direct estimate SE against posterior standard deviations
plot(x=benin_ad1$'SE of Estimate', y=benin_ad1$'F-H SD',
     xlab="Posterior Proportion", ylab="Direct Estimate")
abline(0,1,col="red")

```






**End of report. Code appendix begins on the next page.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document.**