---
title: "CSSS 554: Assignment 2"
subtitle: "Department of Biostatistics @ University of Washington"
# author: 
- Alejandro Hernandez
date: "Winter Quarter 2025"
output: pdf_document
---

```{r setup, include=F}
# Clear environment
rm(list=ls())

# Setup options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, results='hide')
options(knitr.kable.NA = '-', digits = 2)
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "allcode")]

```

```{r load}
# Load relevant packages
library(devtools) 
devtools::install_github("richardli/surveyPrev")
library(surveyPrev) 
library(dplyr) 
library(kableExtra) 
library(labelled) 
library(naniar) 
library(sjlabelled) 
library(haven) 
library(dplyr) 
library(geodata)
library(sf)
library(INLA)
```

1. After creating a DHS account, download and process the required data.

(a) Show the first six lines of the data object created using the getDHSindicator function.

```{r}
indicator <- "unmet_family"
year <- 2010
country <- "Armenia"

## Load data locally
file_path <- "../data/Armenia2016/AMBR72DT/AMBR72FL.DTA" # data file
dhsData <- as.data.frame(read_dta(file_path))
indicator <- "unmet_family"
data <- getDHSindicator(dhsData, indicator = indicator)
head(data)

```

(b) Show the first six lines of the data object created using the adminInfo() function for admin=1.

```{r}
# Get admin 1 info
poly.adm1 <- geodata::gadm(country = country, level = 1, path = tempdir())
poly.adm1 <- sf::st_as_sf(poly.adm1)

adminInfo(poly.adm = poly.adm1, admin = 1, by.adm = "NAME_1")

```

2. Obtain prevalence estimates at admin1, using the surveyPrev::directEST() function. Provide a table with admin names, prevalence estimates, variance of prevalence estimates. [Hint: These are the Hajek estimates.]

```{r}

# Get geo data locally
file_path_geo <- "../data/Armenia2016/AMGE71FL - GPS/AMGE71FL.shp" # shapefile
geo <- st_read(file_path_geo)

## Get prevalence estimates at admin 1 level
# Get cluster information at Admin 1
source("../R/cluster.info.revised.R")
cluster.info <- cluster.info.revised(
  geo = geo, 
  poly.adm1 = poly.adm1, 
  by.adm1 = "NAME_1"
  )

# Estimate prevalence
estimates <- directEST(data = data, cluster.info = cluster.info, admin = 1)

```


3. Using the mapPlot() function, provide two maps, one of the prevalence estimates and the other of the standard errors (which are the square roots of the variances you put in a table in the previous part). And to reiterate, these are at admin1.

```{r}
mapPlot(estimates)
```


**End of report. Code appendix begins on the next page.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document.**