---
title: "CSSS 554: Assignment 2"
subtitle: "Department of Biostatistics @ University of Washington"
author:
- Alejandro Hernandez
date: "Winter Quarter 2025"
output: pdf_document
---

```{r setup, include=F}
# Clear environment
rm(list=ls())

# Setup options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
options(knitr.kable.NA = '-', digits = 2)
labs = knitr::all_labels()
labs = labs[!labs %in% c("setup", "allcode")]

```

```{r load}
# Load relevant packages
# library(devtools) 
# devtools::install_github("richardli/surveyPrev")
library(surveyPrev) 
library(dplyr) 
library(kableExtra) 
library(labelled) 
library(naniar) 
library(sjlabelled) 
library(haven) 
library(dplyr) 
library(geodata)
library(sf)
library(INLA)
library(SUMMER)
library(knitr)

```

1. After creating a DHS account, download and process the required data.

This assignment models the proportion of women with an unmet need for family planning across Admin-1-level regions of Armenia. Armenia is a landlocked country, located in West Asia. It is bordered on the north and east by Georgia and Azerbaijan. On the south and west, it shares borders with Iran and Turkey.

(a) Show the first six lines of the data object created using the getDHSindicator function.

```{r question-1a}
#### QUESTION 1 ####

# Data settings
indicator <- "unmet_family"
year <- 2010
country <- "Armenia"

## Load data locally
file_path <- "../data/Armenia2016/AMBR72DT/AMBR72FL.DTA" # data file
dhsData <- as.data.frame(read_dta(file_path))
indicator <- "unmet_family"
data <- surveyPrev::getDHSindicator(dhsData, indicator = indicator)
# View head
head(data)

```

(b) Show the first six lines of the data object created using the adminInfo() function for admin=1.

```{r question-1b}
# Get admin 1 info
poly.adm1 <- geodata::gadm(country = country, level = 1, path = tempdir())
poly.adm1 <- sf::st_as_sf(poly.adm1)
admin.info <- surveyPrev::adminInfo(poly.adm = poly.adm1, admin = 1, by.adm = "NAME_1")
head(admin.info$data)

```

2. Obtain prevalence estimates at admin1, using the surveyPrev::directEST() function. Provide a table with admin names, prevalence estimates, variance of prevalence estimates. [Hint: These are the Hajek estimates.]

```{r question-2}
#### QUESTION 2 ####

# Get geographic data locally
file_path_geo <- "../data/Armenia2016/AMGE71FL - GPS/AMGE71FL.shp" # shapefile
geo <- st_read(file_path_geo)

## Get prevalence estimates at admin 1 level
# Get cluster information at Admin 1
source("../R/cluster.info.revised.R")
cluster.info <- cluster.info.revised(geo = geo, poly.adm1 = poly.adm1, 
                                     by.adm1 = "NAME_1")

# Estimate prevalence
res_ad1 <- directEST(data = data, cluster.info = cluster.info, admin = 1)
  plot_ad1 <- res_ad1$res.admin1 %>% 
  select("Admin 1 Name"=admin1.name, "Direct Estimate"=direct.est,
         "Variance of Estimate"=direct.var)
plot_ad1

```

3. Using the mapPlot() function, provide two maps, one of the prevalence estimates and the other of the standard errors (which are the square roots of the variances you put in a table in the previous part). And to reiterate, these are at admin1.

```{r question-3}
#### QUESTION 3 ####

# Arrange all estimates for mapPlot
plot_ad1$"SE of Estimate" <- sqrt(plot_ad1$"Variance of Estimate")
plot_ad1$model <- "Percentage of Women with Unmet Family Needs in Armenia"

# Map direct estimates and uncertainty
g1 <- SUMMER::mapPlot(data = plot_ad1, geo = poly.adm1,
              by.data = "Admin 1 Name", by.geo = "NAME_1", is.long = TRUE,
              variable = "model", 
              value = "Direct Estimate", 
              legend.label = "Direct Estimate")
g2 <- mapPlot(data = plot_ad1, geo = poly.adm1,
              by.data = "Admin 1 Name",  by.geo = "NAME_1", is.long = TRUE,
              variable = "model", value = "SE of Estimate", 
              legend.label = "SE of Estimate")
g1
g2

```


**End of report. Code appendix begins on the next page.**

\pagebreak

## Code Appendix

```{r allcode, ref.label = knitr::all_labels(), echo=TRUE, eval=FALSE}
```

**End of document.**